Modelling last glacial cycle ice dynamics in the Alps

### 2018-07-24 articles/cycle/alps-cycle.tex
Apply minor fixes before sending to co-authors.

    The climate forcing driving the ice sheet simulations consist spatially of
    a~present-day monthly mean climatology,
    $\{T_{\mathrm{m}0}, P_{\mathrm{m}0}\}$, modified by temperature lapse-rate
    correction, ${\Delta}T_{\text{LR}}$, temperature
    offset time series, ${\Delta}T_{\text{TS}}$, and time-dependent
    palaeo-precipitation corrections, $\Psi_{\text{PP}}$:
    %
    \begin{align}
      &T_{\mathrm{m}}(t, x, y) = T_{\mathrm{m}0}(x, y) +
                                 {\Delta}T_{\text{LR}}(t, x, y) +
                                 {\Delta}T_{\text{TS}}(t) \,, \\
      &P_{\mathrm{m}}(t, x, y) = P_{\mathrm{m}0}(x, y) \cdot
                                 {\Psi}_{\text{PP}}(t) \,,
    \end{align}
    %
    The present-day monthly mean climatology was bilinearly interpolated from
    near-surface air temperature and precipitation rate fields from
    WorldClim \citep{Hijmans.etal.2005}, representative of the period~1960
    to~1990. Modern climate of the European Alps is characterised by a
    north-south gradient in winter and summer air temperatures
    (Fig.~\ref{fig:inputs}f and g), and an east-west gradient in winter
    precipitation (Fig.~\ref{fig:inputs}h) which is reversed in summer
    (Fig.~\ref{fig:inputs}i). WorldClim data
    were selected as an input to the ice sheet model because they incorporate
    observations from the dense weather station network of central Europe
    \citep[Fig.~1]{Hijmans.etal.2005}. Besides, WorldClim data were
    previously used as climate forcing for PISM to model the LGM extent of the
    former Cordilleran ice sheet in good agreement with geological evidence
    along the southern margin \citep{Seguinot.etal.2014} where weather station
    density is lower than in the Alps. Finally, the last glacial cycle Alpine
    glaciers did not extend over marine areas where WorldClim data are missing.

@@

    Finally, in some simulations (hereafter labelled PP), precipitation was
    reduced with air temperature in
    order to simulate the potential rarefaction of atmospheric moisture in
    colder climates. This was done using an empirical relationship derived from
    observed accumulation rates and oxygen isotopes concentrations in the GRIP
    ice core \citep{Dahl-Jensen.etal.1993},
    %
    \begin{align}
      {\Psi}_{\text{PP}}(t) = \exp[\psi{\Delta}T_{\text{TS}}(t)] \,,
    \end{align}
    %
    with $\psi=0.169/2.4=0.0704$ \citep{Huybrechts.2002}. The control set of
    simulations use constant precipitation, corresponding to $\psi=0$.
    This simple relationship does not reflect the complexity of atmospheric
    circulation changes that governed moisture availability over the Alps
    during the last glacial cycle. Palaeoclimate proxies indicate slightly
    reduced LGM precipitation in western Europe with anomalies diminishing
    eastwards \citep{Wu.etal.2007}. Regional circulation models indicate
    generally dryer conditions during MIS~3 \citep{Barron.Pollard.2002,
    Kjellstrom.etal.2010} but more precipitation south of the Alps during MIS~2
    \citep{Strandberg.etal.2011, Ludwig.etal.2016}. Besides, changes in the
    ice surface topography may have redistributed orographic precipitation.

@@

    In the upper Rhone Valley, observed trimlines are often located near the
    LGM cold-temperate basal thermal transition, or in cold-based areas
    (Fig.~\ref{fig:trimlines}c). The remaining discrepancies may relate to
    temporal migrations of the basal thermal boundary, an absence of sliding in
    warm-based areas, and levelling of small-scale topographic features in the
    1\,km horizontal grid. They call for more detailed comparisons
    spanning the entire Alpine range and specific sensitivity studies to
    relevant basal sliding and ice rheological parameters, and to the uncertain
    subglacial topography. Unfortunately, validation through the bedrock uplift
    rate \citep[cf.][]{Kuchar.etal.2012} is not doable in the Alps due its
    lower values, active tectonics and uncertainties on geological properties
    \citep[cf.][]{Mey.etal.2016}. However, the presence of an upper layer of
    cold ice during the LGM, already found at high altitude in the much warmer
    climate of today \citep[e.g.,][]{Suter.etal.2001, Bohleber.etal.2017}, is
    inevitable \citep[Fig.~3]{Haeberli.Schluchter.1987}.
    In this context, our results challenge the assumption that
    Alpine trimlines mark the maximum upper ice surface elevation of the LGM
    ice cover and call for a more accurate estimation of the thickness of the
    upper layer of cold ice in the Alps.

>>>>

    The climate forcing driving the ice sheet simulations consists spatially of
    a~present-day monthly mean climatology,
    $\{T_{\mathrm{m}0}, P_{\mathrm{m}0}\}$, modified by a temperature lapse-rate
    correction, ${\Delta}T_{\text{LR}}$, temperature
    offset time series, ${\Delta}T_{\text{TS}}$, and time-dependent
    palaeo-precipitation corrections, $\Psi_{\text{PP}}$:
    %
    \begin{align}
      &T_{\mathrm{m}}(t, x, y) = T_{\mathrm{m}0}(x, y) +
                                 {\Delta}T_{\text{LR}}(t, x, y) +
                                 {\Delta}T_{\text{TS}}(t) \,, \\
      &P_{\mathrm{m}}(t, x, y) = P_{\mathrm{m}0}(x, y) \cdot
                                 {\Psi}_{\text{PP}}(t) \,,
    \end{align}
    %
    The present-day monthly mean climatology was bilinearly interpolated from
    near-surface air temperature and precipitation rate fields from
    WorldClim \citep{Hijmans.etal.2005}, representative of the period~1960
    to~1990. Modern climate of the European Alps is characterised by a
    north-south gradient in winter and summer air temperatures
    (Fig.~\ref{fig:inputs}f and g), and an east-west gradient in winter
    precipitation (Fig.~\ref{fig:inputs}h) which is reversed in summer
    (Fig.~\ref{fig:inputs}i). WorldClim data
    were selected as an input to the ice sheet model because they incorporate
    observations from the dense weather station network of central Europe
    \citep[Fig.~1]{Hijmans.etal.2005}. Besides, WorldClim data were
    previously used as climate forcing for PISM to model the LGM extent of the
    former Cordilleran ice sheet in good agreement with geological evidence
    along the southern margin \citep{Seguinot.etal.2014} where weather station
    density is lower than in the Alps. Finally, the last glacial cycle Alpine
    glaciers did not extend over marine areas where WorldClim data are missing.

@@

    Finally, in some simulations (hereafter labelled PP), precipitation was
    reduced with air temperature in
    order to simulate the potential rarefaction of atmospheric moisture in
    colder climates. This was done using an empirical relationship derived from
    observed accumulation rates and oxygen isotopes concentrations in the GRIP
    ice core \citep{Dahl-Jensen.etal.1993},
    %
    \begin{align}
      {\Psi}_{\text{PP}}(t) = \exp[\psi{\Delta}T_{\text{TS}}(t)] \,,
    \end{align}
    %
    with $\psi=0.169/2.4=0.0704$ \citep{Huybrechts.2002}. The control
    simulations use constant precipitation, corresponding to $\psi=0$.
    This simple relationship does not reflect the complexity of atmospheric
    circulation changes that governed moisture availability over the Alps
    during the last glacial cycle. Palaeoclimate proxies indicate slightly
    reduced LGM precipitation in western Europe with anomalies diminishing
    eastwards \citep{Wu.etal.2007}. Regional circulation models indicate
    generally dryer conditions during MIS~3 \citep{Barron.Pollard.2002,
    Kjellstrom.etal.2010} but more precipitation south of the Alps during MIS~2
    \citep{Strandberg.etal.2011, Ludwig.etal.2016}. Besides, changes in the
    ice surface topography may have redistributed orographic precipitation.

@@

    In the upper Rhone Valley, observed trimlines are often located near the
    LGM cold-temperate basal thermal transition or within cold-based areas
    (Fig.~\ref{fig:trimlines}c). The remaining discrepancies between the
    trimlines locations and the modelled basal thermal boundary may relate to
    temporal migrations of the basal thermal boundary, an absence of sliding in
    warm-based areas, and levelling of small-scale topographic features in the
    1\,km horizontal grid. They call for more detailed comparisons
    spanning the entire Alpine range and specific sensitivity studies to
    relevant basal sliding and ice rheological parameters, and to the uncertain
    subglacial topography. Unfortunately, validation through the bedrock uplift
    rate \citep[cf.][]{Kuchar.etal.2012} is not doable in the Alps due its
    lower values, active tectonics and uncertainties on geological properties
    \citep[cf.][]{Mey.etal.2016}. However, the presence of an upper layer of
    cold ice during the LGM, already found at high altitude in the much warmer
    climate of today \citep[e.g.,][]{Suter.etal.2001, Bohleber.etal.2017}, is
    inevitable \citep[Fig.~3]{Haeberli.Schluchter.1987}.
    In this context, our results challenge the assumption that
    Alpine trimlines mark the maximum upper ice surface elevation of the LGM
    ice cover and call for a more accurate estimation of the thickness of the
    upper layer of cold ice in the Alps.
